---
title: "Did OTB (Over the Board) Chess Recover After COVID-19?"
subtitle: "Evidence from FIDE Standard rating lists, 2015-2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(broom)

players <- readRDS("data/processed/players.rds")
period_data <- players %>%
  filter(!is.na(period)) %>%
  mutate(log_games = log1p(games))
covid_window <- as.Date(c("2020-03-01", "2021-12-31"))
```

## The Question

**Did over-the-board (OTB) competitive chess fully recover after COVID-19?**

- Online chess boomed during COVID (Queen's Gambit, Lichess/chess.com surge)
- FIDE Standard ratings reflect **OTB tournament games only**
- Did the real-world tournament ecosystem bounce back — or stay smaller?

## The Dataset

- **FIDE Standard rating lists** — 7 monthly snapshots, 2015-2026
- Each snapshot's `games` field = rated OTB games played **in the prior
  month**
- Active rated player = `rating > 0` and not flagged inactive

| Snapshot | Games reflect | Period |
|---|---|---|
| 2015-02 | Jan 2015 | (inflation only) |
| 2019-01 | Dec 2018 | Pre-COVID |
| 2020-01 | Dec 2019 | Pre-COVID |
| 2020-07 | Jun 2020 | COVID |
| 2021-01 | Dec 2020 | COVID |
| 2022-01 | Dec 2021 | Post-COVID |
| 2026-06 | May 2026 | Post-COVID |

## Active Players Over Time

```{r}
#| label: active-plot
#| fig-width: 9
#| fig-height: 5

active_counts <- players %>%
  count(snapshot, ref_month, name = "active_players") %>%
  arrange(ref_month)

ggplot(active_counts, aes(ref_month, active_players)) +
  annotate("rect", xmin = covid_window[1], xmax = covid_window[2],
           ymin = -Inf, ymax = Inf, fill = "grey85") +
  geom_line(color = "#5b9fff", linewidth = 1.2) +
  geom_point(color = "#f0c040", size = 3) +
  labs(x = "Reference month (games played in)", y = "Active players") +
  theme_minimal(base_size = 16)
```

## Did Players Keep Playing? (Participation Rate)

```{r}
#| label: participation-plot
#| fig-width: 9
#| fig-height: 5

participation <- players %>%
  group_by(snapshot, ref_month) %>%
  summarise(pct_played = 100 * mean(played), .groups = "drop") %>%
  arrange(ref_month)

ggplot(participation, aes(ref_month, pct_played)) +
  annotate("rect", xmin = covid_window[1], xmax = covid_window[2],
           ymin = -Inf, ymax = Inf, fill = "grey85") +
  geom_line(color = "#51cf66", linewidth = 1.2) +
  geom_point(color = "#f0c040", size = 3) +
  labs(x = "Reference month", y = "% of active players who played >=1 rated game") +
  theme_minimal(base_size = 16)
```

## Did Participation Recover? (Chi-Squared Test)

```{r}
#| label: chisq

contingency <- table(period_data$period, period_data$played)
contingency_pct <- round(100 * prop.table(contingency, margin = 1), 1)
chisq_result <- chisq.test(contingency)
cramers_v <- sqrt(unname(chisq_result$statistic) / sum(contingency))
```

`period (3 levels) x played (yes/no)`, all active players in the 6
Pre/COVID/Post-COVID snapshots:

- Pre-COVID: `r contingency_pct["Pre-COVID","TRUE"]`% played
- COVID: `r contingency_pct["COVID","TRUE"]`% played
- Post-COVID: `r contingency_pct["Post-COVID","TRUE"]`% played

chi-squared(`r unname(chisq_result$parameter)`) = `r format(round(unname(chisq_result$statistic), 1), big.mark = ",", scientific = FALSE)`,
p `r format.pval(chisq_result$p.value, eps = .001)`, Cramer's V = `r round(cramers_v, 3)`

::: {.notes}
With N in the millions, p-values are essentially always significant —
Cramer's V and the raw percentages are the figures that matter.
:::

## Games-Played Intensity by Period (ANOVA + Tukey)

```{r}
#| label: anova

aov_fit <- aov(log_games ~ period, data = period_data)
aov_table <- summary(aov_fit)[[1]]
eta_sq <- aov_table["period", "Sum Sq"] / sum(aov_table[["Sum Sq"]])
tukey_df <- as.data.frame(TukeyHSD(aov_fit)$period)
```

DV: `log1p(games)` (log-transformed — `games` is heavily right-skewed and
zero-inflated).

- F(`r aov_table["period","Df"]`, `r format(aov_table["Residuals","Df"], big.mark = ",", scientific = FALSE)`) = `r format(round(aov_table["period","F value"], 1), big.mark = ",", scientific = FALSE)`,
  p `r format.pval(aov_table["period","Pr(>F)"], eps = .001)`
- eta-squared = `r round(eta_sq, 4)`
- Tukey HSD, **Post-COVID vs Pre-COVID**: diff (log scale) =
  `r round(tukey_df["Post-COVID-Pre-COVID","diff"], 3)`,
  p `r format.pval(tukey_df["Post-COVID-Pre-COVID","p adj"], eps = .001)`

## Controlling for Age & Title (Regression)

```{r}
#| label: regression

reg_data <- period_data %>% filter(!is.na(age))
lm_fit <- lm(log_games ~ period + age + title_has, data = reg_data)
lm_summary <- broom::tidy(lm_fit) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
r2 <- summary(lm_fit)$r.squared
```

`log1p(games) ~ period + age + title_has`, n =
`r format(nobs(lm_fit), big.mark = ",")`, R-squared = `r round(r2, 3)`

```{r}
#| label: regression-table
knitr::kable(lm_summary)
```

::: {.notes}
Compare the period coefficients here to the raw chi-squared/ANOVA
differences — do they survive controlling for age and title composition?
:::

## The Twist: Rating Inflation, 2015 vs 2026

```{r}
#| label: inflation
#| fig-width: 8
#| fig-height: 4.5

inflation_data <- players %>% filter(snapshot %in% c("2015-02", "2026-06"))
inflation_summary <- inflation_data %>%
  group_by(snapshot) %>%
  summarise(n = n(), mean_rtg = mean(rating), sd_rtg = sd(rating), .groups = "drop")
sd_ratio <- max(inflation_summary$sd_rtg) / min(inflation_summary$sd_rtg)
t_result <- t.test(rating ~ snapshot, data = inflation_data, var.equal = (sd_ratio < 1.2))

ggplot(inflation_data, aes(rating, fill = snapshot)) +
  geom_histogram(alpha = 0.5, binwidth = 25, position = "identity") +
  scale_fill_manual(values = c("2015-02" = "#5b9fff", "2026-06" = "#f0c040")) +
  labs(x = "Rating", y = "Number of players", fill = NULL) +
  theme_minimal(base_size = 16)
```

Mean rating: 2015 = `r round(inflation_summary$mean_rtg[inflation_summary$snapshot=="2015-02"], 1)`,
2026 = `r round(inflation_summary$mean_rtg[inflation_summary$snapshot=="2026-06"], 1)`
(`r ifelse(sd_ratio < 1.2, "pooled-variance", "Welch")` t-test,
p `r format.pval(t_result$p.value, eps = .001)`)

::: {.notes}
Ties back to the main story: if participation recovered/grew, did the
rating pool itself shift (younger/newer player influx, FIDE rating-floor
policy changes)?
:::

## Conclusion

- Participation rate dropped sharply during COVID (32.1% -> 3.9%) and
  **only partially recovered** by 2026 (24.7%) — still ~7 points below
  the Pre-COVID level
- Effect sizes (Cramer's V = 0.29, eta-squared = 0.07) are **moderate**
  despite very small p-values — large N inflates significance
- The period pattern **holds** after controlling for age and title
  composition (Post-COVID coefficient -0.073 vs. raw Tukey diff -0.071 —
  essentially unchanged)
- Bonus: the rating pool shows **lower** mean ratings in 2026 vs 2015
  (1736.9 vs 1799.3) — likely diluted by the influx of new, lower-rated
  players (active pool nearly doubled, 115.5K -> 217.5K)

::: {.notes}
Caveat: the same player ID can appear across multiple snapshots, which
violates the independence assumption for chi-squared/ANOVA/regression.
This is standard for repeated cross-sections of administrative data and is
acknowledged here rather than corrected with out-of-scope panel methods.
:::
