---
title: "Did OTB (Over the Board) Chess Recover After COVID-19?"
subtitle: "Evidence from FIDE Standard rating lists, 2015-2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(broom)

players <- readRDS("data/processed/players.rds")
period_data <- players %>%
  filter(!is.na(period)) %>%
  mutate(log_games = log1p(games))
covid_window <- as.Date(c("2020-03-01", "2021-12-31"))
```

## The Question

**Did over-the-board (OTB) competitive chess fully recover after COVID-19?**

- Online chess boomed during COVID (Queen's Gambit, Lichess/chess.com surge)
- FIDE Standard ratings reflect **OTB tournament games only**
- Did the real-world tournament ecosystem bounce back — or stay smaller?

## The Dataset

- **FIDE Standard rating lists** — 7 monthly snapshots, 2015-2026
- Each snapshot's `games` field = rated OTB games played **in the prior
  month**
- Active rated player = `rating > 0` and not flagged inactive

| Snapshot | Games reflect | Period |
|---|---|---|
| 2015-02 | Jan 2015 | (inflation only) |
| 2019-01 | Dec 2018 | Pre-COVID |
| 2020-01 | Dec 2019 | Pre-COVID |
| 2020-07 | Jun 2020 | COVID |
| 2021-01 | Dec 2020 | COVID |
| 2022-01 | Dec 2021 | Post-COVID |
| 2026-06 | May 2026 | Post-COVID |

## Active Players Over Time

```{r}
#| label: active-plot
#| fig-width: 9
#| fig-height: 5

active_counts <- players %>%
  count(snapshot, ref_month, name = "active_players") %>%
  arrange(ref_month)

ggplot(active_counts, aes(ref_month, active_players)) +
  annotate("rect", xmin = covid_window[1], xmax = covid_window[2],
           ymin = -Inf, ymax = Inf, fill = "grey85") +
  geom_line(color = "#5b9fff", linewidth = 1.2) +
  geom_point(color = "#f0c040", size = 3) +
  labs(x = "Reference month (games played in)", y = "Active players") +
  theme_minimal(base_size = 16)
```

## Did Players Keep Playing? (Participation Rate)

```{r}
#| label: participation-plot
#| fig-width: 9
#| fig-height: 5

participation <- players %>%
  group_by(snapshot, ref_month) %>%
  summarise(pct_played = 100 * mean(played), .groups = "drop") %>%
  arrange(ref_month)

ggplot(participation, aes(ref_month, pct_played)) +
  annotate("rect", xmin = covid_window[1], xmax = covid_window[2],
           ymin = -Inf, ymax = Inf, fill = "grey85") +
  geom_line(color = "#51cf66", linewidth = 1.2) +
  geom_point(color = "#f0c040", size = 3) +
  labs(x = "Reference month", y = "% of active players who played >=1 rated game") +
  theme_minimal(base_size = 16)
```
